# Food groups and periodontal disease {#food-groups}
Chapter compiled **`r Sys.Date()`**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(here)
library(tidyverse)
```

The previous chapter detailed treelet transformation of NHANES dietary data and associations with periodontal outcomes. Whilst many of the nutrient groupings identified by the TCs were sensible, it was not possible to determine which types of *diet* provided these nutrients. This chapter focuses on associations between diet (rather than nutrient intake) and periodontal disease in an effort to extract real-world dietary patterns that could form the basis for diet recommendations to improve periodontal outcomes.

## Dietary data
Data were drawn from NHANES waves F to H (2009/10 to 2013/14) and the same process of exclusions used to generate the analysis cohort. The NHANES dietary data contains files detailing the individual foods and drinks consumed during the two 24 hour survey periods for each individual. The number of different food items consumed across the cohort is very large and contains many items consumed by very few people, thus providing little information on which to base inference. Therefore, food items were classified into food groups defined in the USDA Food and Nutrient Database for Dietary Studies, version 5 [@ahuja_usda_2012]. When a food group was consumed in both 24 hour periods, the average intake value was used. When a group was only consumed in a single period, this value was used (where the food group was not consumed in either period, the value of zero was used).

```{r prepare-modelling-dataset, include=FALSE}
source(here("Code", "diet_periodontal.R"))
source(here("Code", "helper_functions.R"))
```

Food groups were selected for inclusion in the treelet analysis such that they were non-overlapping and gave the maximum resolution possible.

```{r table-of-food-groups}
fgrp %>% 
  select(grp_code, grp_description) %>% 
  distinct() %>%
  transmute(`Food group` = grp_code, 
            `Description` = grp_description, 
            `In dendrogram` = if_else(grp_code %in% names(food_groups_nhanes), "Yes", "")) %>% 
  kable(caption="USDA food groups included in treelet analysis.")

```

An important modelling decision is how to adjust for total energy intake when determining dietary patterns. The initial approach (as in the nutrient intake chapter) was to use correlations of raw intake quantities (g/day) as the input to the Treelet algorithm and then adjust for total energy intake (`KCAL`) in regression modelling of the outcome. There was no association between `KCAL` and CAL outcomes indicating that the observed effects were largely independent of energy intake.

However, when the median intakes of food groups were compared by TC score it became apparent that some of the TCs were highly correlated with overall quantity of food intake. For example, higher intake of almost all food groups was found in  high deciles of the first treelet component. This may have occured simply because when consuming more, the probability of including a greater variety of items (at reasonable quantities) increases. Consuming more of everything is not a sensible dietary recommendation so the problem remains how to separate the dietary patterns from the overall quantities consumed.

The approach chosen was to scale the intakes of each food group by the total energy consumption for that individual to give an indication of the proportion of the diet attributable to that food group. This was done crudely by dividing the grammes intake by total `KCAL` however this does not account for the differences in energy values between foods and may downweight the foods never consumed in large quantities.

When comparing the initial raw quantity analysis with an analysis of variables scaled by `KCAL` the cluster trees were similar but the variables included in each TC and the loadings were not. `KCAL` was also excluded from the regression models to avoid overadjustment. When plotting the resulting TC deciles against unadjusted daily intake (g) the summaries no longer give the impression that simply consuming more is better. 

```{r treelet-analysis-food-groups, include = F}
source(here("Code", "food_groups.R"))
```

The full height dendrogram for the included food groups revealed plausible associations among food groups. For example, the most closely correlated groups were lettuce and salad dressing, followed by milk and ready-to-eat cereals.

```{r dendrogram, fig.height = 11, fig.width=7, fig.cap="Dendrogram of food group intake scaled by overall energy intake."}
ggdendrogram(food_groups_dendro, rotate = T)
```

```{r full-treelet-scree-plot}
# Scree plot to choose number of treelet components to retain
food_groups_tc_full$tcv %>% 
  ggplot(aes(x = Component, y = Variance)) +
  geom_line() +
  geom_point() +
  theme_light() + 
  ggtitle("Scree plot for treelet tranformation of food group intake data")
```

The scree plot indicates that it would be appropriate to retain 8 components, as beyond this point there is little incremental increase in variance explained with each additional component.


The cross validation plot indicates that a cut level of 28 would be appropriate, as there is little additional increase in score with increasing cut level beyond this point.

```{r treelet-cross-validation-plot, echo=F}
apply(cvs, 1, mean) %>% 
  enframe() %>%
  transmute(Cross_validation_score = value, Cut_level = as.numeric(str_extract(name, "[0-9]{1,2}"))) %>% 
  ggplot(aes(x = Cut_level, y = Cross_validation_score)) +
  geom_line() +
  theme_light() + 
  ggtitle("Cross validation performance of treelet transformation") +
  labs(x = "Cut level", y = "Cross validation score")

```
  
TC1, TC3 and TC5 showed relatively broad distribution of scores. The remaining TCs had large spikes at the lower end of each distribution. These are likely to represent individuals that did not consume any of the food groups forming that TC. 

```{r TC-score-distributions}
# Distribution of TC scores
food_groups_tc_scores %>% 
  as_tibble() %>% 
  gather() %>% 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50) + 
  facet_wrap(~key, scales="free_x") + 
  theme_light() +
  labs(x = "TC score", y = "Frequency") + 
  ggtitle("Distribution of Treelet Component scores for food group analysis")
```

There was very little correlation among the TCs (Kendalls's tau was used due to the presence of ties).

```{r TC-score-correlations}
kable(food_groups_tc_cor, caption = "Correlation of TC scores")
```

Components TC1 was broadly loaded but the others were sparse indicating relatively discrete patterns of association among food groups.

```{r TC-loading-distribution}
# Number of non-zero loadings for each TC
food_groups_loadings %>% 
  group_by(Component) %>% 
  summarise(`Non-zero scores` = sum(abs(Value) > 0)) %>% 
  kable(caption = "Summary of loading patterns for Treelet components in food group analysis.")
```

TC1 characterised diets high in salads, vegetables, poultry, seafood, fruits (but not fruit juices) and water or tea to drink. TC2 loaded on milk, ready-to-eat cereals and bananas (the breakfast TC). TC3 loaded on bread, luncheon meats and table fats, coffee and sugar (the coffee and sandwich TC). TC4 loaded on fruit juice drinks and fruit flavoured drinks (but not pure fruit juices). TC5 loaded on the same food groups as TC3 but represented a contrast between low intake of bread, meat and table fats and high intake of coffee and sugar (the coffee and no sandwich TC). TC6 loaded on beef, potatoes and beans and corn (the Mexican TC?). TC7 loaded on candy, crackers, pretzels and low calorie carbonated soft drinks (the snack TC). TC8 loaded on pancakes, eggs and French toast (the egg TC).

```{r treelet-loading-plots, fig.height=11, fig.width=7}
ggplot(food_groups_loadings, aes(x = Variable, y = Value)) +
  facet_wrap(~Component, nrow = 1) +
  geom_bar(stat = "identity") + 
  scale_y_continuous(breaks = c(0, 0.5)) +
  coord_flip() + 
  theme_light() +
  ggtitle("TC loading plot for analysis of food group intake")
```


## Clinical attachment loss (CAL)
### Beta regression modelling

Associations between TCs and CAL were assessed using a series of Beta regressions, as for nutrient intake. These included linear functions of each TC, quadratic functions and a model splitting each TC into deciles to capture non-linearities in potential associations with CAL. The most parsimonious (using BIC) included the adjustment variables and quadratic fits for the eight TCs in both the mean and precision models.

```{r food-group-CAL-beta-regression}
summary(fg4)
```

There was strong evidence that the majority of both linear and quadratic terms for TCs were associated with CAL at $P<0.05$, indicating curvilinear relationships between TCs and CAL. However, many of the estimates were of small magnitude. To interpret the quadratic fit, predictions of the conditional effect of each TC were made for male, non-diabetic, non-smokers of average age and with average energy intake and for whom all but the focal TC were set at mean values. 
 
The predicted range of CAL response was calculated for each TC to enable comparison of the magnitude of each association. Prediction ranges were restricted to the between the 10th and 90th percentiles because at the extremes of the distribution back transformation from the logistic scale can lead to inflated slope estimates. 

```{r TC-predictions-from-beta-regression}
cal_pred_frame_grps %>% 
  select(component, qtile) %>% 
  cbind(CAL_pred = predict(fg4, newdata = cal_pred_frame_grps, type = "response")) %>% 
ggplot(aes(x = qtile, y = CAL_pred)) +
  geom_line() +
  facet_wrap(~component) +
  theme_light() +
  labs(y = expression("Proportion of sites with CAL">="3mm"), x = "Percentile of TC distribution") +
  ggtitle("Predicted extent of clinical attachment loss")
```

Increasing TC1 and TC7 scores were associated with substantial decreases in proportion of sites with CAL. TC5 was associated with moderate increase in CAL across the range of observed scores. All other TCs showed less predicted variation in CAL.

```{r CAL-predict-polynomial}
cal_pred_frame_grps %>% 
  select(component, qtile) %>% 
  cbind(CAL_pred = predict(fg4, newdata = cal_pred_frame_grps, type = "response")) %>% 
  filter(qtile >= 10, qtile <= 90) %>% 
  group_by(Component = component) %>% 
  summarise(`Prediction range (%)` = round((max(CAL_pred) -min(CAL_pred))*100, 1)) %>% 
  kable(caption ="Prediction range of clinical attachment loss by treelet component.")

```

### Robust regresssion modelling

Whilst capturing the underlying associations with the TCs, beta regression models systematically overestimated the proportion of sites with CAL, with poor coverage of the tails of the distribution. A number of alternative models were tested to improve model fit. 

```{r comparison-of-actual-and-predicted,include=FALSE}
nh_grps %>% transmute(Measurement = "Actual", value = prop_CAL_sites3mm_beta) %>% 
  bind_rows(enframe(predict(fg4)) %>% mutate(Measurement = "Predicted: beta regression (quadratic)")) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(aes(fill = Measurement), bins = 50, position = "dodge") +
  theme_light() +
   labs(y = expression("Proportion of sites with CAL">="3mm"), x = "Count")
```

Standard logistic regression of the number of sites with CAL, weighted by the number of sites assessed was fitted. Instead of polynomial functions, each TC was classified into deciles. Comparison with a similarly specified beta regression accounting for heteroscedasticity, the logistic model was less conservative in terms of the relative risk estimates. 

```{r CAL-logistic-regression}
# Compare standard and beta regression parameters
enframe(formatC(exp(coef(lg1)), format = "f", digits = 2), value = "Logistic regression") %>% 
  inner_join(enframe(formatC(exp(coef(fg6)), format = "f", digits = 2), value = "Beta regression"), by = "name") %>% 
  rename(Term = name) %>% 
  kable(caption = "Comparison of Odds ratios estimated using logistic and beta regression.")
```
   
Predictions from the logisitic model was also more closely grouped towards the lower range of the data. The model also systematically underestimated the actual mean.

```{r comparison-logistic-vs-beta}
nh_grps %>% transmute(Measurement = "Actual", value = prop_CAL_sites3mm_beta) %>% 
  bind_rows(enframe(predict(fg6)) %>% mutate(Measurement = "Predicted: beta regression (decile)"),
            enframe(predict(lg1, type = "response")) %>% mutate(Measurement = "Predicted: logistic regression (decile)")) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(aes(fill = Measurement), bins = 50, position = "dodge") +
  theme_light() +
   labs(y = expression("Proportion of sites with CAL">="3mm"), x = "Count")
```

Another aspect of both the beta and logistic models is that they are models of the *mean* of the distribution. The right skewed nature of the proportion of sites with CAL indicates that the *median* might be a more meaningful statistic to model

```{r CAL-by-TC-scores}
nh_grps %>% 
  select(prop_CAL_sites3mm, matches("_dec")) %>% 
  gather(component, TC_decile, -prop_CAL_sites3mm) %>%
  mutate(TC_decile = factor(TC_decile, levels = 1:10)) %>% 
  filter(component == "TC1_dec") %>% 
  ggplot(aes(x = TC_decile, y = prop_CAL_sites3mm)) +
   geom_boxplot(outlier.shape = NA, fill = NA, notch = T) +
  geom_point(data = TC1_means, mapping = aes(y = value, colour = Measurement)) +
   theme_light() +
  labs(y = expression("Proportion of sites with CAL">="3mm"), x = "Decile of TC distribution") +
  ggtitle("Extent of clinical attachment loss")
```

Standard quantile regression using the `quantreg` package [@R-quantreg] was considered but predictions are not constrained to the interval [0,1] so is not appropriate for modelling proportions. Instead, robust logistic quantile regression was performed using the `lqr` package [@R-lqr]. These models are constrained to the [0,1] interval.

The robust quantile model was a much better fit to the data than the mean models with the predicted medians closely matching actual medians. Quantile regression on the lower quartile was also reasonably accurate but regression on the upper quartile tended to overestimate this value.

```{r robust-quantile-histogram}
nh_grps %>% transmute(Measurement = "Actual", value = prop_CAL_sites3mm_beta) %>% 
  bind_rows(invlogit(rob1$fitted.values) %>% 
              enframe() %>% 
              mutate(Measurement = "Predicted: robust quantile (decile)")) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(aes(fill = Measurement), bins = 50, position = "dodge") +
  theme_light() +
   labs(y = expression("Proportion of sites with CAL">="3mm"), x = "Count")
```


```{r CAL-median-by-TC-scores}
nh_grps %>% 
  select(prop_CAL_sites3mm, matches("_dec")) %>% 
  gather(component, TC_decile, -prop_CAL_sites3mm) %>%
  mutate(TC_decile = factor(TC_decile, levels = 1:10)) %>% 
  filter(component == "TC1_dec") %>% 
  ggplot(aes(x = TC_decile, y = prop_CAL_sites3mm)) +
   geom_boxplot(outlier.shape = NA, fill = NA, notch = T) +
  geom_point(data = TC1_medians, mapping = aes(y = value, colour = Measurement)) +
   theme_light() +
  labs(y = expression("Proportion of sites with CAL">="3mm"), x = "Decile of TC distribution") +
  ggtitle("Extent of clinical attachment loss")
```

A slightly more flexible quantile regression fitting was attempted using the `cdfquantreg` package [@R-cdfquantreg] which allows joint modelling of both the median and dispersion functions (analagous to beta regression). This model did produce predictions spanning the entire (0,1) interval but tended to overpredict in both tails of the distribution. Therefore, robust quantile regression was selected to model associations between TC scores and CAL.


```{r cdf-quantile-histogram}
nh_grps %>% transmute(Measurement = "Actual", value = prop_CAL_sites3mm_beta) %>% 
  bind_rows(predict(lrq1) %>% 
              enframe() %>% 
              mutate(Measurement = "Predicted: CDF quantile")) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(aes(fill = Measurement), bins = 50, position = "dodge") +
  theme_light() +
   labs(y = expression("Proportion of sites with CAL">="3mm"), x = "Count")
```


```{r CAL-cdf-median-by-TC-scores}
nh_grps %>% 
  select(prop_CAL_sites3mm, matches("_dec")) %>% 
  gather(component, TC_decile, -prop_CAL_sites3mm) %>%
  mutate(TC_decile = factor(TC_decile, levels = 1:10)) %>% 
  filter(component == "TC1_dec") %>% 
  ggplot(aes(x = TC_decile, y = prop_CAL_sites3mm)) +
   geom_boxplot(outlier.shape = NA, fill = NA, notch = T) +
  geom_point(data = TC1_cdf_medians, mapping = aes(y = value, colour = Measurement)) +
   theme_light() +
  labs(y = expression("Proportion of sites with CAL">="3mm"), x = "Decile of TC distribution") +
  ggtitle("Extent of clinical attachment loss")
```

There was strong evidence from the robust quantile regresssion that proportion of sites with CAL decreased with increasing TC1. As for the beta regressions, increasing age was associated with increasing CAL, as was smoking, diabetes and male gender. There was also evidence that CAL decreased in deciles 5 and above of TC7. Other TCs were not associated with CAL.

Age was the greatest source of variation in proportion of sites with CAL. The conditional estimates indicated that variation in CAL with TC1 was considerable, of similar magnitude to that associated with smoking status and gender and approximately twice the variation associated with diabetes status. Variation in CAL with TC7 was slightly less than TC1 but still comparable with smoking and gender.


```{r robust-coefficients}
kable(rob1_coef, caption = "Estimates from robust logistic quantile regression on CAL. Coefficients and odds ratios  given.")

```

Combining the conditional estimates to make predictions for some representative groups, the decrease in median proportion of sites with CAL between TC1 decile 1 (`r round(tc1_dec1, 2)`%) and TC1 decile 10 (`r round(tc1_dec10, 2)`%) was equivalent to the change expected with decrease in age from 59 to 45 years. For TC7, the decrease from decile 1 (`r round(tc7_dec1, 2)`%) to decile 10 (`r round(tc7_dec10)`%) was equivalent to a decrease in age from 59 to 47 years. It was notable that the decrease in CAL only occured in deciles five and above of TC7.


```{r medians-other-variables, include = FALSE}
nh_grps %>% 
  group_by(diabetes) %>% 
  summarise(CAL_median = median(prop_CAL_sites3mm))

nh_grps %>% 
  group_by(RIAGENDR) %>% 
  summarise(CAL_median = median(prop_CAL_sites3mm))

nh_grps %>% 
  group_by(SMQ020) %>% 
  summarise(CAL_median = median(prop_CAL_sites3mm))

```

## Cohort characteristics by treelet component scores   
We summarised the characteristics of cohort members by deciles of TC1 and TC7 to determine whether these diets are associated with particular demographic groups. 

There was a clear association between the gender and TC1 with males overrepresented in deciles 1 to 5 and females overrepresented in deciles 6 to 10 (almost 70% in decile 10 were female). Likelihood of ever having smoked decreased with increasing TC1 decile. There was no consistent association between diabetes and TC1 and no substantial variation in age among TC1 deciles.

```{r characteristics-by-TC1-quantile}
nh_grps %>% 
inner_join(nhanes %>% 
             select(SEQN, Age = RIDAGEYR), by = "SEQN") %>% 
    group_by(`TC1 decile` = TC1_dec) %>% 
  summarise(Age = paste0(round(mean(Age)), " (", round(sd(Age), digits = 1), ")"),
            Male = perc(sum(RIAGENDR == 1)/length(RIAGENDR)),
            Diabetes = perc(sum(diabetes)/length(diabetes)),
            `Ever smoked` = perc(sum(SMQ020 != 2)/length(SMQ020))) %>% 
  knitr::kable(caption = "Distribution of individual characteristics by TC1 decile.")
```

There was no evidence of associations between TC7 and age, diabetes or smoking status. There was some indication that males were overrepresented in deciles 1 to 5 and females in 6 to 10 but this pattern was not as strong as that observed for TC1.

```{r characteristics-by-TC7-quantile}
nh_grps %>% 
inner_join(nhanes %>% 
             select(SEQN, Age = RIDAGEYR), by = "SEQN") %>% 
  group_by(`TC7 decile` = TC7_dec) %>% 
  summarise(Age = paste0(round(mean(Age)), " (", round(sd(Age), digits = 1), ")"),
            Male = perc(sum(RIAGENDR == 1)/length(RIAGENDR)),
            Diabetes = perc(sum(diabetes)/length(diabetes)),
            `Ever smoked` = perc(sum(SMQ020 != 2)/length(SMQ020))) %>% 
  knitr::kable(caption = "Distribution of individual characteristics by TC7 decile.")

```

In summary, both TC1 and TC7 captured variation between male and female diets, although it should be noted that the regression models quantified variation in CAL by the TCs independent of gender and the other adjustment variables.

## Diet profiles by treelet component scores    
To gain insight into the magnitude of the dietary variation captured by the TC scores, average daily consumption (g/day) was calculated by TC decile for each food group with a non-zero loading for each TC. These plots highlight two aspects of the TCs, the proportion of people in each decile consuming any of the food groups on the survey dates and the median quantity consumed among those that did consume. 


For TC1 the most commonly consumed items were water, tomatoes (`VEG4`) and 'other vegetables' (`VEG8`). The largest gradients in proportion consumed were found for lettuce (`VEG5`) and salad dressings (`FAT2`).

```{r TC1-food-group-daily-intake, fig.cap = "Daily intake of food groups by TC1 decile."}
ggplot(ExtractTCIntake(daily_intake, 1, TC1_dec), aes(y = Variable, x = TC1_dec)) +
  geom_tile(aes(fill = any_intake)) +
  geom_text(aes(label = grms_intake), colour = "white") +
  labs(x = "TC1 decile", y = "Food group") +
  guides(fill = guide_colourbar(title = "Proportion with any intake")) +
  theme_minimal()
```

The greatest variation in the proportion of individuals consuming a food group for TC7 was for `GRAIN5` (crackers, popcorn, pretzels, corn chips). The sample was split 60/40 into those that consumed these items and those that did not. There was a also a strong gradient of increased quantities consumed, with the median in decile 10 more than five times that in decile 5. Half of the sample consumed no or negligible quantities of low calorie carbonated soft drinks. There was a more than four-fold increase in median quantity consumed from decile 5 to decile 10. A similar pattern was observed for candy (`SUGAR2`) although the quantities consumed were much smaller.

```{r TC7-food-group-daily-intake, fig.cap = "Daily intake of food groups by TC7 decile.", warning=FALSE}
ggplot(ExtractTCIntake(daily_intake, 7, TC7_dec), aes(y = Variable, x = TC7_dec)) +
  geom_tile(aes(fill = any_intake)) +
  geom_text(aes(label = grms_intake), colour = "white") +
  labs(x = "TC7 decile", y = "Food group") +
  guides(fill = guide_colourbar(title = "Proportion with any intake")) +
  theme_minimal()
```

Finding an association between hight TC7 scores and reduced risk of CAL was surprising given that the three food groups involved are not considered healthy and the sugars in candy might be expected to have a direct detrimental effect on oral health. A possible hypothesis explaining the observed patterns is that reduced CAL risk was driven by food groups *not* consumed in the higher TC7 deciles but not highly correlated with any of the groups within TC7. Drinks were considered as a candidate and the strongest correlation among all drink groups was the negative one between `BEV242` (in TC7) and `BEV241` (regular carbonated soft drinks). Other correlations with regular carbonated drinks include with regular fruit juices and fruit flavoured drinks `BEV231` (-ve) and with water (-ve).
   

```{r correlations-among-drinks}
food_groups_nhanes %>% 
    select(matches("BEV|WATER")) %>% 
  cor() %>% 
  # cor(method = "kendall") %>% 
  kable(caption = "Correlation among drink variables.", digits = 2)
```

Daily intake of regular carbonated soft drinks was added to the daily intake figure and there appears to be some evidence that low calorie drinks replaced regular drinks in the higher TC7 deciles. The proportion of individuals consuming any regular carbonated drinks decreased from around 40% in deciles 1-8 to 23% in decile 10. Among those with any intake, the quantities consumed remained similar (~1 can per day).

```{r drink-replacement-figure, warning=F}
daily_intake %>% 
  # Select only those food groups with a non zero score for the TC of interest
  inner_join(food_groups_loadings %>% 
               filter(Component == "TC7", Value != 0 | Variable == "BEV241") %>%  
             mutate(grp_code = as.character(Variable)),
             by = "grp_code") %>% 
  # Calculate summary statistics for each decile and food group
  group_by(TC7_dec, Variable) %>% 
  # Proportion with any intake 
   summarise(any_intake = round(sum(grms>0)/length(grms), digits = 2),
             # Median intake among those with any intake
             grms_intake = round(median(grms[grms>0]))) %>% 
  ungroup() %>% 

  ggplot(aes(y = Variable, x = TC7_dec)) +
  geom_tile(aes(fill = any_intake)) +
  geom_text(aes(label = grms_intake), colour = "white") +
  labs(x = "TC7 decile", y = "Food group") +
  guides(fill = guide_colourbar(title = "Proportion with any intake")) +
  theme_minimal()

```


```{r nutrient-correlations-TC1}
kable(cor_tc1, caption = "Correlations between TC1 and nutrient variables.", digits = 2)
```

```{r nutrient-correlations-TC7}
kable(cor_tc7, caption = "Correlations between TC7 and nutrient variables.", digits = 2)
```


## Nutrient profiles by treelet component scores
The final stage of the analysis was to examine the distribution of nutrient intake by TC1 and TC7. The aim was to investigate whether nutrients associated with periodontal disease in other studies showed substantial gradients in the diets associated with CAL in this study.  


```{r nutrients-by-TC1, fig.width=8, fig.height = 12}
nh_grps %>% select(SEQN, TC1_dec) %>% 
  inner_join(dietary, by = "SEQN") %>% 
  gather("Nutrient", "intake", -SEQN, -TC1_dec) %>% 
  group_by(TC1_dec, Nutrient) %>% 
  # Proportion with any intake 
   summarise(any_intake = round(sum(intake>0)/length(intake), digits = 2),
             # Median intake among those with any intake
             grms_intake = round(median(intake[intake>0]))) %>% 
  ungroup() %>% 
  
  # Select only nutrients with strong correlations 
  inner_join(cor_tc1, by = "Nutrient") %>%
  arrange(r) %>% 
  mutate(Nutrient = as_factor(Nutrient)) %>% 

  # Plot 
  ggplot(aes(y = Nutrient, x = TC1_dec)) +
  geom_tile(aes(fill = any_intake)) +
  geom_text(aes(label = grms_intake), colour = "white") +
  labs(x = "TC1 decile") +
  guides(fill = guide_colourbar(title = "Proportion with any intake")) +
  theme_minimal()

```

```{r nutrients-by-TC7, fig.width=8, fig.height = 12}
nh_grps %>% select(SEQN, TC7_dec) %>% 
  inner_join(dietary, by = "SEQN") %>% 
  gather("Nutrient", "intake", -SEQN, -TC7_dec) %>% 
  group_by(TC7_dec, Nutrient) %>% 
  # Proportion with any intake 
   summarise(any_intake = round(sum(intake>0)/length(intake), digits = 2),
             # Median intake among those with any intake
             grms_intake = round(median(intake[intake>0]))) %>% 
  ungroup() %>% 

  # Select only nutrients with strong correlations 
  inner_join(cor_tc7, by = "Nutrient") %>%
  arrange(r) %>% 
  mutate(Nutrient = as_factor(Nutrient)) %>% 

  ggplot(aes(y = Nutrient, x = TC7_dec)) +
  geom_tile(aes(fill = any_intake)) +
  geom_text(aes(label = grms_intake), colour = "white") +
  labs(x = "TC7 decile") +
  guides(fill = guide_colourbar(title = "Proportion with any intake")) +
  theme_minimal()
```

### Intake of sugary drinks

A potential explanation for the association between TC7 and CAL is that those with a high TC7 score have less exposure to carboyhdrates (mainly sugars) in drinks and that this is distinct from overall carbohydrate intake. The quantities of carbohydrates and sugars consumed as drinks were calculated for each individual and plotted by TC7 decile.

```{r carb-food-vs-drink}
food_drink_intake %>% 
  select(TC7_dec, Intake_Carbs_Drink, Intake_Carbs_Food) %>% 
  gather("Variable", "Intake" , -TC7_dec) %>% 
  ggplot(aes(x = TC7_dec, y = Intake)) +
  geom_boxplot() +
  #geom_violin() +
  facet_wrap(~Variable, scales = "free") +
  theme_light()

```

There was evidence of a weak negative correlation between TC7 score and carbohydrate intake from drinks (r = `r round(cor(food_drink_intake$TC7, food_drink_intake$Intake_Carbs_Drink), 2)`) but not with intake from food (r = `r round(cor(food_drink_intake$TC7, food_drink_intake$Intake_Carbs_Food), 2)`). When added to the CAL model, carbohydrate intake from drinks was associated with increased proportion of sites with CAL but intake from food was not. Including these variables attenuated slightly the assocation with TC7. The same was observed for TC1 indicating that intake of carbohydrates from drinks is a component of this dietary pattern also. This distinction between food and drink sources of carbohydrates has not been observed with regard to periodontal outcomes before.

```{r CAL-food-drink-carbs-coefs}
kable(rob2_coef, caption = "Estimates from robust logistic quantile regression on CAL. Coefficients and odds ratios  given.")
```


