# Figures and tables for publication {#paper}

```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
library(here)
library(checkpoint)
checkpoint_date <- "2019-07-31"
checkpoint(checkpoint_date, use.knitr = TRUE, project = here())
library(tidyverse)
```

Chapter compiled **`r Sys.Date()`**   

R checkpoint **`r checkpoint_date`**   


This chapter comprises the figures and tables describing these analyses as submitted to the _American Journal for Clinical Nutrition_.

```{r prepare-modelling-dataset, include=FALSE}
source(here("Code", "diet_periodontal.R"))
source(here("Code", "helper_functions.R"))
```

```{r run-analyses, include=FALSE}
source(here("Code", "treelet_food_groups.R"))
source(here("Code", "food_groups.R"))
```

## Figures for text
NHANES respondents 2009-2014: `r nrow(filter(nhanes_all, between(SDDSRVYR, 6, 8)))`   
Persons aged $\geq30$ years: `r nrow(filter(nhanes_all, 
                                    between(SDDSRVYR, 6, 8),
                                    RIDAGEYR >= 30))`   
Persons aged $\geq30$ years with medical examination:`r nrow(filter(nhanes_all, 
                                    between(SDDSRVYR, 6, 8),
                                    RIDAGEYR >= 30,
                                    RIDSTATR == 2))`   
Persons aged $\geq30$ years with medical examination who are dentate:`r nrow(filter(nhanes_all, 
                                    between(SDDSRVYR, 6, 8),
                                    RIDAGEYR >= 30,
                                    RIDSTATR == 2,
                                    tooth_count > 0))`   
Persons aged $\geq30$ years with medical examination who are dentate and eligible for periodontal exam:
                            `r nrow(filter(nhanes_all, 
                                    between(SDDSRVYR, 6, 8),
                                    RIDAGEYR >= 30,
                                    RIDSTATR == 2,
                                    tooth_count > 0,
                                    EXCLU != 1))`   
Persons aged $\geq30$ years with medical examination who are dentate, eligible and had a completed periodontal exam:
                            `r nrow(filter(nhanes_all, 
                                    between(SDDSRVYR, 6, 8),
                                    RIDAGEYR >= 30,
                                    RIDSTATR == 2,
                                    tooth_count > 0,
                                    EXCLU != 1,
                                    PDSTS == 1))`   
Persons aged $\geq30$ years with medical examination who are dentate, eligible and had a completed periodontal exam with $\geq1$ sites assessed for clinical attachment loss:
                            `r nrow(filter(nhanes_all, 
                                    between(SDDSRVYR, 6, 8),
                                    RIDAGEYR >= 30,
                                    RIDSTATR == 2,
                                    tooth_count > 0,
                                    EXCLU != 1,
                                    PDSTS == 1,
                                    CAL_sites_assessed > 0))`   
Persons aged $\geq30$ years with medical examination who are dentate, eligible and had a completed periodontal exam with $\geq1$ sites assessed for clinical attachment loss and had valid dietary data:
                            `r nrow(filter(nhanes_all, 
                                    between(SDDSRVYR, 6, 8),
                                    RIDAGEYR >= 30,
                                    RIDSTATR == 2,
                                    tooth_count > 0,
                                    EXCLU != 1,
                                    PDSTS == 1,
                                    CAL_sites_assessed > 0,
                                    PD_sites_assessed > 0,
                                    dietary == T))`   

Mean age of the cohort: `r round(mean(nhanes$RIDAGEYR))`   
SD age of the cohort: `r round(sd(nhanes$RIDAGEYR), 1)`   
Percentage male: `r perc(sum(nhanes$RIAGENDR == 1)/length(nhanes$RIAGENDR))`   
Percentage with diabetes: `r perc(sum(nhanes$diabetes)/length(nhanes$diabetes))`   
Percentage ever smoked: `r perc(sum(nhanes$SMQ020 == 1)/length(nhanes$RIAGENDR))`   
25th, 50th and 75th percentiles of tooth count: `r quantile(nhanes$tooth_count, probs = c(0.25, 0.5, 0.75))`   
25th, 50th and 75th percentiles of proportion of sites with CAL $\geq3$mm: `r perc(quantile(nhanes$prop_CAL_sites3mm, probs = c(0.25, 0.5, 0.75)))`   
Strongest correlation among TC scores: $\tau=$ `r sort(na.omit(as.numeric(food_groups_tc_cor)))[1]`   
Correlation between intake of low calorie and regular carbonated drinks: r = `r round(food_groups_cor["BEV242", "BEV241"], digits = 2)`      
Correlation between intake of Theobromine and TC7 score: `r round(cor_tc7["THEO"], digits = 2)`   
Predicted CAL extent for male, non-smoker, non-diabetic in decile 5 for each TC, aged 46: `r round(age46, 1)`   
Predicted CAL extent for male, non-smoker, non-diabetic in decile 5 for each TC, aged 61: `r round(age61, 1)`    
Percentage of the sample with <20 teeth: `r round(prop.table(table(nhanes$tooth_count<20))[["TRUE"]]*100, 1)`   

Percentage of cohort with only one recorded 24 hour recall: `r perc(sum(food_total_grms_per_day$DRDINT == 1)/nrow(food_total_grms_per_day))`    

Mean CAL at site level: `r round(mean(perio_cal_site$value), 1)`
Mean of proportion of sites with CAL $\geq3$mm: `r mean(nhanes$prop_CAL_sites3mm)`
Proportion of individuals with at least one site with CAL $\geq3$mm: `r round(sum(nhanes$CAL_sites3mm > 0)/nrow(nhanes)*100)`
Mean PPD at site level: `r round(mean(perio_pocket_site$value), 1)`
Mean of proportion of sites with PPD $\geq4$mm: `r mean(nhanes$prop_PD_sites4mm)`
Proportion of individuals with at least one site with PPD $\geq4$mm: `r round(sum(nhanes$PD_sites4mm > 0)/nrow(nhanes)*100)`


## Tables

```{r food-groups-included}
fgrp %>% 
  select(grp_code, grp_description) %>% 
  distinct() %>%
  filter(grp_code %in% names(food_groups_nhanes)) %>% 
  transmute(`Food group` = grp_code, 
            `Description` = grp_description) %>% 
  arrange(`Food group`) %>% 
  kable(caption="USDA food groups included in treelet analysis.")

```

```{r selected-robust-coefficients}
rob1_coef %>% 
  mutate(Variable = case_when(Term == "RIDAGEYR" ~ "Age",
                              Term == "RIAGENDR" ~ "Female",
                              Term == "SMQ020" ~ "Never smoked",
                              Term == "diabetesTRUE" ~ "Diabetes",
                              str_detect(Term, "(TC1|TC7)_dec") ~ str_replace(Term, "_dec", " Decile "))) %>% 
  filter(!is.na(Variable)) %>% 
  transmute(Variable,         
            `Odds Ratio` = FormatOddsRatio(Estimate, `Std. Error`),
            P = paste(format.pval(`Pr(>|z|)`, digits = 1, eps = 0.001), Sig)) %>% 
  
  # Add the predicted CAL medians
  full_join(TC1_medians %>% 
              mutate(TC = "TC1") %>% 
              rbind(TC7_medians %>% 
                      mutate(TC = "TC7")) %>% 
              filter(str_detect(Measurement, "robust")) %>% 
              mutate(Variable = paste(TC, "Decile", TC_decile),
                        `Median sites with CAL` = perc(value))
                    , by = "Variable") %>% 
  
  # Add other periodontal measures
  left_join(bind_rows(TC1_perio, TC7_perio), by = "Variable") %>% 
  
  mutate_at(vars(`Median sites with CAL`, P), ~if_else(is.na(.), "", .)) %>% 
  mutate(`Odds Ratio` = if_else(is.na(`Odds Ratio`), "1.00", `Odds Ratio`)) %>% 
  arrange(TC, TC_decile) %>% 
  select(Variable, `Odds Ratio`, P, 
         `Median sites with CAL`, `Mean CAL (mm)`, `Mean PPD (mm)`) %>%   
kable(caption = "Association between dietary Treelet Component scores, adjustment variables and extent of clinical attachment loss.")

```

```{r nutrients-by-TC1}
nh_grps %>% select(SEQN, TC1_dec) %>% 
  inner_join(dietary, by = "SEQN") %>% 
  gather("Nutrient", "intake", -SEQN, -TC1_dec) %>% 
  group_by(TC1_dec, Nutrient) %>% 
     # Median intake among those with any intake
   summarise(grms_intake = round(median(intake[intake>0]))) %>% 
  ungroup() %>% 
  
  # Select only nutrients with strong correlations 
  inner_join(cor_tc1 %>% 
               enframe(name = "Nutrient", value = "tau") %>% 
  filter(abs(tau)>0.15)
  , by = "Nutrient") %>%
  mutate(tau =  round(tau, 2)) %>% 
  
  # Label nutrient column
  inner_join(diet_names, by = c("Nutrient" = "Variable Name")) %>% 
  
  spread(TC1_dec, grms_intake) %>%
  mutate(TC = 1) %>% 
  select(TC, Nutrient = Description, tau, matches("[0-9]{1}")) %>% 

  arrange(desc(tau)) %>% 

  bind_rows(
    nh_grps %>% select(SEQN, TC7_dec) %>% 
  inner_join(dietary, by = "SEQN") %>% 
  gather("Nutrient", "intake", -SEQN, -TC7_dec) %>% 
  group_by(TC7_dec, Nutrient) %>% 
     # Median intake among those with any intake
   summarise(grms_intake = round(median(intake[intake>0]))) %>% 
  ungroup() %>% 
  
  # Select only nutrients with strong correlations 
  inner_join(cor_tc7 %>% 
               enframe(name = "Nutrient", value = "tau") %>% 
  filter(abs(tau)>0.15)
  , by = "Nutrient") %>%
  mutate(tau =  round(tau, 2)) %>% 
  
  # Label nutrient column
  inner_join(diet_names, by = c("Nutrient" = "Variable Name")) %>% 
  
  spread(TC7_dec, grms_intake) %>%
    mutate(TC = 7) %>% 
  select(TC, Nutrient = Description, tau, matches("[0-9]{1}")) %>% 
    

  arrange(desc(tau)) 
  ) %>% 
  
    kable(caption = "Median daily intake of nutrients by TC decile and correlation with TC.")
```
    
    
## Supplementary tables

```{r food-drink-model}
rob2_coef %>% 
  mutate(Variable = case_when(Term == "RIDAGEYR" ~ "Age",
                              Term == "RIAGENDR" ~ "Female",
                              Term == "SMQ020" ~ "Never smoked",
                              Term == "diabetesTRUE" ~ "Diabetes",
                              Term == "Intake_Carbs_Drink" ~ "Carbohydrate intake - drinks",
                              Term == "Intake_Carbs_Food" ~ "Carbohydrate intake - food",
                              str_detect(Term, "(TC1|TC7)_dec") ~ str_replace(Term, "_dec", " Decile "))) %>% 
   filter(!is.na(Variable)) %>% 
  transmute(Variable,         
            `Odds Ratio` = FormatOddsRatio(Estimate, `Std. Error`),
            P = paste(format.pval(`Pr(>|z|)`, digits = 1, eps = 0.001), Sig)) %>% 
  
  # Add the predicted CAL medians
  full_join(TC1_medians_rob2 %>% 
              mutate(TC = "TC1") %>% 
              rbind(TC7_medians_rob2 %>% 
                      mutate(TC = "TC7")) %>% 
              filter(str_detect(Measurement, "robust")) %>% 
              mutate(Variable = paste(TC, "Decile", TC_decile),
                        `Median sites with CAL` = perc(value))
                    , by = "Variable") %>% 
  
  
  mutate_at(vars(`Median sites with CAL`, P), ~if_else(is.na(.), "", .)) %>% 
  mutate(`Odds Ratio` = if_else(is.na(`Odds Ratio`), "1.00", `Odds Ratio`)) %>% 
  arrange(TC, TC_decile) %>% 
  select(Variable, `Odds Ratio`, P, 
         `Median sites with CAL`) %>% 
kable(caption = "Association between dietary Treelet Component scores, adjustment variables and extent of clinical attachment loss.")
```

```{r selected-robust-coefficients-multivariate-nutrient-density}
rob4_coef %>% 
  mutate(Variable = case_when(Term == "RIDAGEYR" ~ "Age",
                              Term == "RIAGENDR" ~ "Female",
                              Term == "SMQ020" ~ "Never smoked",
                              Term == "diabetesTRUE" ~ "Diabetes",
                              str_detect(Term, "(TC1|TC7)_dec") ~ str_replace(Term, "_dec", " Decile "))) %>% 
  filter(!is.na(Variable)) %>% 
  transmute(Variable,         
            `Odds Ratio` = FormatOddsRatio(Estimate, `Std. Error`),
            P = paste(format.pval(`Pr(>|z|)`, digits = 1, eps = 0.001), Sig)) %>% 
  
  mutate(`Odds Ratio` = if_else(is.na(`Odds Ratio`), "1.00", `Odds Ratio`)) %>% 
  #arrange(TC, TC_decile) %>% 
  select(Variable, `Odds Ratio`, P) %>%   
kable(caption = "Association between dietary Treelet Component scores, adjustment variables and extent of clinical attachment loss. Additional adjustment for total nutrient intake using the multivariate nutrient density model.")

```



## Figures
```{r dendrogram, fig.height = 11, fig.width=7, fig.cap="Dendrogram of food group intake produced using treelet transformation."}
ggdendrogram(food_groups_dendro, rotate = T) +
  geom_hline(linetype = 2, yintercept = 28) +
  ggsave(here("Papers", "revision", "fig2.tiff"), device = "tiff", dpi = 300, width = 7, height = 8, units = "in", compression = "lzw", type = "cairo")
```

```{r treelet-loading-plots, fig.height=11, fig.width=7}
ggplot(food_groups_loadings, aes(x = Variable, y = Value)) +
  facet_wrap(~Component, nrow = 1) +
  geom_bar(stat = "identity") + 
  scale_y_continuous(breaks = c(0, 0.5)) +
  labs(y = "Component loading", x = "Food Group") +
  coord_flip() + 
  theme_light() +
  ggsave(here("Papers", "revision", "fig3.tiff"), device = "tiff", dpi = "print", width = 7, height = 11, units = "in", compression = "lzw", type = "cairo")

```

```{r TC1-food-group-daily-intake, fig.height=5, fig.width=5.7, fig.cap = "Daily intake (g/day) of food groups by TC1 decile. Median g/day among those with any intake.", message=FALSE}
ExtractTCIntake(daily_intake, 1, TC1_dec) %>% 
  mutate(text_col = if_else(any_intake > 0.5, "white", "black")) %>% 
ggplot(aes(y = Variable, x = TC1_dec)) +
  geom_tile(aes(fill = any_intake)) +
  scale_fill_gradient(high = "black", low = "white", breaks = c(0, 0.2, 0.4, 0.6, 0.8,1), limits = c(0,1)) +
   geom_text(aes(label = grms_intake, colour = text_col)) +
  scale_colour_identity()+
  labs(x = "TC1 decile", y = "Food group") +
  guides(fill = guide_colourbar(title = "Proportion with any intake")) +
  theme_minimal() +
  theme(legend.position = "top") +
   ggsave(here("Papers", "revision", "fig4.tiff"), device = "tiff", dpi = "print", width = 5.7, height = 5, units = "in", compression = "lzw", type = "cairo")

```

```{r TC7-food-group-daily-intake, fig.height=2, fig.width=5.7, fig.cap = "Daily intake of food groups by TC7 decile. Median g/day among those with any intake.", warning=FALSE, message = FALSE}
ExtractTCIntake(daily_intake, 7, TC7_dec) %>% 
  mutate(text_col = if_else(any_intake > 0.5, "white", "black")) %>% 
    filter(TC7_dec %in% c("1","2","3") == F) %>% 
ggplot(aes(y = Variable, x = TC7_dec)) +
  geom_tile(aes(fill = any_intake)) +
   scale_fill_gradient(high = "black", low = "white", breaks = c(0, 0.2, 0.4, 0.6, 0.8,1), limits = c(0,1)) +
  geom_text(aes(label = grms_intake, colour = text_col)) +
  scale_colour_identity() +
  labs(x = "TC7 decile", y = "Food group") +
  guides(fill = guide_colourbar(title = "Proportion with any intake")) +
  theme_minimal() +
  theme(legend.position = "top") +
    ggsave(here("Papers", "revision", "fig5.tiff"), device = "tiff", dpi = "print", width = 5.7, height = 2, units = "in", compression = "lzw", type = "cairo")

```

```{r CAL-age, fig.cap = "Mean CAL (site level) by age."}
perio_cal_site %>% 
  group_by(RIDAGEYR) %>% 
  summarise(`Mean CAL` = mean(value)) %>% 
  ggplot(aes(x = RIDAGEYR, y =  `Mean CAL`)) + 
  geom_line()
```

```{r PPD-age, fig.cap = "Mean pocket depth (site level) by age."}
perio_pocket_site %>% 
  group_by(RIDAGEYR) %>% 
  summarise(`Mean CAL` = mean(value)) %>% 
  ggplot(aes(x = RIDAGEYR, y =  `Mean CAL`)) + 
  geom_line()
```

