# Adjusting for socio-economic status and ethnicity {#SES}

```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
library(here)
library(checkpoint)
checkpoint_date <- "2020-04-02"
checkpoint(checkpoint_date, use.knitr = TRUE, project = here())
library(tidyverse)
library(kableExtra)
```

Chapter compiled **`r Sys.Date()`**   

R checkpoint **`r checkpoint_date`**   

This chapter comprises additional analyses following the third round of reviewing at the _American Journal for Clinical Nutrition_. Updated figures and tables are also provided.

```{r prepare-modelling-dataset, include=FALSE}
source(here("Code", "diet_periodontal.R"))
source(here("Code", "helper_functions.R"))
```

```{r run-analyses, include=FALSE}
source(here("Code", "treelet_food_groups.R"))
source(here("Code", "food_groups.R"))
source(here("Code", "food_groups_PCA.R"))
```


The editors requested that all models be refitted to include measures of socio-economic status (SES) and ethnicity. In line with previous studies of periodontal disease using NHANES data, ethnicity `RIDRETH1` and highest educational attainment `DMDEDUC2` were selected. Ethnicity was included because it is plausible that it may influence both access to dental care and biological processes (e.g. inflammation) involved in periodontitis. Educational attainment was selected for SES because it can be considered a lifecourse indicator  and is therefore relevant for assessing associations with a chronic condition such as periodontitis. More volatile commonly used measures of SES including annual family income INDFMIN` were considered but not selected because these may change rapidly and also because this referred to the family rather than the individual, which was the unit of analysis.   

Measures of access to dental care drawn from the oral health questionnaire `OHQ` were considered as they represent a plausible causal pathway between SES and periodontitis. The variable `OHQ030` which indicates time since last dental visit was considered but discounted because the reasons for a recent visit (detailed in `OHQ033`) were a mixture of standard check-ups (perhaps indicating better oral health behaviour) and symptom driven attendance. Another question covering the reasons why individuals did not access dental care despite need `OHQ770` and `OHQ780A` were also discounted because it is likely to relate more closely to acute dental problems rather than chronic conditions. Patient judgement of need for periodontal care, especially in mild cases is likely to be a poor. Also, these questions were only asked in waves G and H of NHANES and so the size of the dataset would have been reduced by a third.


## Figures for text

Predicted CAL extent for white, male, high school graduate, non-smoker, non-diabetic in decile 5 for each TC, aged 48: `r round(rob10_age48, 1)`   
Predicted CAL extent for white, male, high school graduate, non-smoker, non-diabetic in decile 5 for each TC, aged 62: `r round(rob10_age62, 1)`  


AIC of robust logistic quantile regression of median CAL proportion with three category smoking status and TCs categorised into deciles: `r round(rob10$AIC)`
AIC of robust logistic quantile regression of median CAL proportion with three category smoking status and straight line relationships only with TCs: `r round(rob11$AIC)`

```{r selected robust-coefficients-SES}
rob10_coef %>% 
  kable(caption = "Association between dietary Treelet Component scores, adjustment variables and extent of clinical attachment loss.")

```


## Tables

```{r robust-coefficients-SES}
rob10_summary %>% 
  filter(str_detect(Metric, "Mean", negate = T)) %>% 
  kable(caption = "Association between dietary Treelet Component scores and extent of clinical attachment loss (CAL).") #%>% 
  # scroll_box(width = "100%")
```

```{r descriptive-stats-by-CAL-quartile-SES}
bind_rows(
nhanes %>% 
  rename(`Mean Age (SD)` = RIDAGEYR) %>% 
  SummariseContinuousTwoway(vars(`Mean Age (SD)`), CAL_sites_quartile),
nhanes %>%
  mutate(Sex = factor(if_else(RIAGENDR == 1, "Male", "Female"), levels = c("Male", "Female")),
         Diabetes = factor(if_else(diabetes, "Diabetes", "No diabetes"), levels = c("No diabetes", "Diabetes")),
         Smoking = factor(case_when(smoking == "Never" ~ "Never smoked",
                             smoking == "Former" ~ "Former smoker",
                             smoking == "Current" ~ "Current smoker"), 
                          levels = c("Never smoked", "Former smoker", "Current smoker")),
         Ethnicity = factor(case_when(RIDRETH1 == 1 ~ "Mexican American",
                                      RIDRETH1 == 2 ~ "Other Hispanic",
                                      RIDRETH1 == 3 ~ "Non-Hispanic White",
                                      RIDRETH1 == 4 ~ "Non-Hispanic Black",
                                      RIDRETH1 == 5 ~ "Other race - including Multi-Racial"),
                            levels = c("Mexican American",
                                       "Other Hispanic",
                                       "Non-Hispanic White",
                                       "Non-Hispanic Black",
                                       "Other race - including Multi-Racial")),
         `Education level` = factor(case_when(DMDEDUC2 == 1 ~ "Education:Less than 9th grade",
                                       DMDEDUC2 == 2 ~ "Education:9-11th grade",
                                       DMDEDUC2 ==3 ~ "Education:High school graduate",
                                       DMDEDUC2 == 4 ~ "Education:Some college",
                                       DMDEDUC2 == 5 ~ "Education:College graduate or above",
                                       DMDEDUC2 == "(Missing)" ~ "Education:Refused/don't know"),
         levels = c("Education:Less than 9th grade", 
                    "Education:9-11th grade", 
                    "Education:High school graduate", 
                    "Education:Some college", 
                    "Education:College graduate or above",
                    "Education:Refused/don't know")))  %>% 
  SummariseDiscreteTwoway(vars(Sex, Diabetes, Smoking, Ethnicity, `Education level`), CAL_sites_quartile)
) %>% 
  mutate(Level = coalesce(Level, Variable)) %>% 
  select(Level, one_of(as.character(1:4))) %>% 
  kable(caption = "Cohort characteristics by quartile of CAL extent.")

```

## Supplementary tables

```{r robust-coefficients-PCA-SES}
rob10_summary_pca %>% 
  filter(str_detect(Metric, "Mean", negate = T)) %>% 
  kable(caption = "Association between dietary Principal Component scores and extent of clinical attachment loss (CAL).") #%>% 
  # scroll_box(width = "100%")
```
