---
title: "Food groups and periodontal disease"
author: "David Wright"
date: "4 April 2019"
output: html_document
---

```{r setup, include=FALSE}
library(checkpoint)
checkpoint("2019-03-11")

library(knitr)
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(betareg) # For Beta regression
library(Formula) # For multipart formulas used in Beta regression
library(lmtest) # For likelihood ratio tests of Beta regressions
library(splines)
library(broom)
```

These notes describe analysis of associations between food groups and periodontal outcomes. Data were drawn from NHANES waves F to H (2009/10 to 2013/14). 

## Cohort construction

```{r prepare modelling dataset, include=FALSE}
source(here("Code", "diet_periodontal.R"))
source(here("Code", "helper_functions.R"))
```

Selection of food groups to include in the treelet analysis is described below.

```{r table of food groups included}
# Select combination of non-overlapping groups that gives the maximum resolution possible
food_groups_nhanes <- food_grps_grms_per_day %>% 
  
  select(SEQN, 
         #matches("BEV(11|21|22|231|232|233|241|242)$"),
         matches("BEV(0|21|22|231|232|233|241|242)$"),
         EGG0,
         matches("FAT(1|2)$"),
         matches("FRUIT(11|2|31|32|33|34|35)$"),
         matches("GRAIN(1|21|22|23|3|4|5|6)$"),
         matches("MEAT(1|2|3|4|5|6|7|8)$"),
         matches("MILK(1|2|3|4)$"),
         matches("SUGAR(1|2)$"),
         matches("VEG(1|2|3|4|5|6|7|8)$"),
         WATER1)


fgrp %>% 
  select(grp_code, grp_description) %>% 
  distinct() %>%
  transmute(`Food group` = grp_code, 
            `Description` = grp_description, 
            `In dendrogram` = if_else(grp_code %in% names(food_groups_nhanes), "Yes", "")) %>% 
  kable()

```



```{r run treelets on food groups, include = F}

# Scale data so can intepret treelets on correlation scale
food_groups_scl <- food_groups_nhanes %>%
  select(-SEQN) %>%
  scale()

food_groups_cor <- cor(food_groups_scl)

# Generate maximum height tree
# Save basis vectors at all cut levels so the most useful can be selected later
food_groups_tree_full <- Run_JTree(food_groups_cor, maxlev = ncol(food_groups_cor)-1, whichsave = 1:(ncol(food_groups_cor)-1))
# Extract the treelet components and associated variances
food_groups_tc_full <- TTVariances(food_groups_tree_full, food_groups_cor)

# Dendrogram of maximum height tree
food_groups_dendro <- dendro_data(ConvertTTDendro(food_groups_tree_full))

```

The full height dendrogram for the included food groups is given below.
```{r dendrogram, fig.height = 11, fig.width=7, echo = F}
ggdendrogram(food_groups_dendro, rotate = T)
```


```{r full treelet scree plot, echo=F}
# Scree plot to choose number of treelet components to retain
food_groups_tc_full$tcv %>% 
  ggplot(aes(x = Component, y = Variance)) +
  geom_line() +
  geom_point() +
  theme_light() + 
  ggtitle("Scree plot for treelet tranformation of food group intake data")
```

The scree plot indicates that it would be appropriate to retain 8 components, as beyond this point there is little incremental increase in variance explained with each additional component.

```{r treelet cross validation, include =F}
## Initial analysis - data driven selection of number of components and cut points

## Cross validation ##
# Data driven cross validation to choose a cut level that can describe the 
# data with few components (TCs)
# Set the desired number of components (m) based on scree plot and then find the best cut level
# Cross validation scores for each cut level
m <- 8
cvs <- replicate(5, CrossValidateTT(food_groups_scl, m = m))

```

```{r treelet cross validation plot, echo=F}
apply(cvs, 1, mean) %>% 
  enframe() %>%
  transmute(Cross_validation_score = value, Cut_level = as.numeric(str_extract(name, "[0-9]{1,2}"))) %>% 
  ggplot(aes(x = Cut_level, y = Cross_validation_score)) +
  geom_line() +
  theme_light() + 
  ggtitle("Cross validation performance of treelet transformation") +
  labs(x = "Cut level", y = "Cross validation score")

```
    
The cross validation plot indicates that a cut level of 25 would be appropriate, as there is little additional increase in score with increasing cut level beyond this point.

```{r fit reduced treele, include = F}

# Project selected components to get scores for each individual using the selected cut level
# Cut level selected based on  cross validation and inspection of dendrogram

# Original
food_groups_tc_reduced <- TTVariances(food_groups_tree_full, cor(food_groups_scl), cut_level = 25, components = m)
food_groups_tc_scores <- food_groups_scl %*% food_groups_tc_reduced$tc

### Add TC scores to main NHANES dataset
nh <- bind_cols(nhanes, as_tibble(food_groups_tc_scores)) %>% 
  # Classify TC scores into deciles
  mutate_at(vars(matches("^TC[0-9]{1,2}$")), list(~as.factor(ntile(., n = 10)))) %>% 
  rename_at(vars(matches("^TC[0-9]{1,2}$")), ~paste0(., "_dec")) %>% 
  # Raw TC scores
  bind_cols(as_tibble(food_groups_tc_scores)) %>% 
  inner_join(dietary, by = "SEQN")

#food_groups_out <- bind_cols(diet, data.frame(food_groups_tc_scores))

# Extract loadings for TCs
food_groups_loadings <- food_groups_tc_reduced$tc %>% 
  as_tibble(rownames = "Variable") %>% 
  gather(Component, Value, -Variable) %>% 
  # Order by leaf labelling order
  mutate(Variable = factor(Variable, levels = as.character(food_groups_dendro$labels$label)))

```

There was very little correlation among the TCs.

```{r TC score distributions, echo=F}
# Distribution of TC scores
food_groups_tc_scores %>% 
  as_tibble() %>% 
  gather() %>% 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50) + 
  facet_wrap(~key, scales="free_x") + 
  theme_light() +
  labs(x = "TC score", y = "Frequency") + 
  ggtitle("Distribution of Treelet Component scores")

# Correlation of scores
# fairly low correlation across scores
food_groups_tc_cor <- cor(food_groups_tc_scores) %>% 
  formatC(digits = 2, format = "f")
food_groups_tc_cor[upper.tri(food_groups_tc_cor, diag = T)] <- ""
kable(food_groups_tc_cor, caption = "Correlation of TC scores")

```

Components TC1 and TC2 were broadly loaded but the others were sparse.


```{r TC loading distribution, echo = F}
# Number of non-zero loadings for each TC
food_groups_loadings %>% 
  group_by(Component) %>% 
  summarise(`Non-zero scores` = sum(abs(Value) > 0)) %>% 
  kable()


```

```{r treelet loading plots, fig.height=11, fig.width=7}
ggplot(food_groups_loadings, aes(x = Variable, y = Value)) +
  facet_wrap(~Component, nrow = 1) +
  geom_bar(stat = "identity") + 
  coord_flip() + 
  theme_light() +
  ggtitle("TC loading plot for analysis of food group intake")

```


## Clinical attachment loss (CAL)

Associations between TCs and CAL were assessed using Beta regression.

```{r food group CAL beta regression, echo=F}
nh$prop_CAL_sites3mm_beta <- PropTransform(nh$prop_CAL_sites3mm)

fg1 <- betareg(formula = prop_CAL_sites3mm_beta ~ RIDAGEYR + RIAGENDR + 
     SMQ020 + diabetes + KCAL + TC1 + TC2 + TC3 + TC4 + TC5 + TC6 + 
     TC7 + TC8 | RIDAGEYR + RIAGENDR + SMQ020 + diabetes, data = nh)
summary(fg1)

```

High scores for TC1 associated with lower proportion of sites with CAL. High scores for TC5 and TC6 were associated with a greater proportion of sites with CAL. Gradients in risk across the range of each TC were in the region of 5-6% (TC1) and 3-4% (TC5 and TC6).

```{r TC predictions from beta regression, echo =F}
# Model to predict from
mpred <- fg1

# Take quantiles and stack
tc_qtx <- StackQuantiles(food_groups_tc_scores, probs = seq(0.02, 0.98, by = 0.01))

# Prediction for average aged, male, non-smoker, non-diabetic, average energy intake
cal_pred_frame <- cbind(RIDAGEYR = mean(nh$RIDAGEYR), 
       RIAGENDR = 1, 
       SMQ020 = 2, 
       diabetes = F,
       KCAL = mean(nh$KCAL),
       tc_qtx) %>% 
  as_tibble()
  

cal_pred_frame %>% 
  select(component, qtile) %>% 
  cbind(CAL_pred = predict(mpred, newdata = cal_pred_frame, type = "response")) %>% 
ggplot(aes(x = qtile, y = CAL_pred)) +
  geom_line() +
  facet_wrap(~component) +
  theme_light() +
  labs(y = expression("Proportion of sites with CAL">="3mm"), x = "Percentile of TC distribution") +
  ggtitle("Predicted extent of clinical attachment loss")


```



```{r CAL TC polynomial models, include = F}
# Cubic spline fit
# cal5 <- update(cal4, .~. - TC1 - TC2 - TC3 - TC4 - TC5 - TC6 - TC7 + bs(TC1) + bs(TC2) + bs(TC3) + bs(TC4) + bs(TC5) + bs(TC6) + bs(TC7)|., data = nh)
# summary(cal5)
# # Quadratic fit
# cal6 <- update(cal4, .~. + I(TC1^2) + I(TC2^2) + I(TC3^2) + I(TC4^2) + I(TC5^2) + I(TC6^2) + I(TC7^2) |., data = nh)
# summary(cal6)
```
